---
title: "Backend provenance and CUDA diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backend provenance and CUDA diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`cudacellr` combines sparse CPU preprocessing with CUDA-aware PCA and nearest
neighbours. A function accepting `device = "cuda"` is therefore not
necessarily an end-to-end GPU workflow. Version 0.2.0 records each actual
stage so hybrid execution remains visible.

The main tutorial below is a complete CPU workflow. The later CUDA section is
optional and executes only when diagnostics prove that a usable device exists.

## A reproducible CPU workflow

`cudacellr` accepts features in rows and cells in columns. Start with a small
sparse count matrix:

```{r cpu-data}
library(cudacellr)
library(Matrix)

set.seed(42)
counts <- matrix(
  rpois(60 * 30, lambda = 2),
  nrow = 60,
  ncol = 30,
  dimnames = list(
    paste0("gene_", seq_len(60)),
    paste0("cell_", seq_len(30))
  )
)
counts[1, ] <- counts[1, ] + 1
counts <- Matrix(counts, sparse = TRUE)
```

Every stage is independently callable:

```{r cpu-stages}
normalized <- cuda_normalize_counts(counts)
variable <- cuda_hvg(normalized, n_top = 20)
pca <- cuda_cell_pca(
  counts,
  n_hvg = 20,
  n_components = 5,
  device = "cpu"
)
neighbors <- cuda_cell_neighbors(
  pca,
  k = 5,
  batch_size = 8,
  device = "cpu"
)
```

The convenience workflow produces the same sequence with one call:

```{r cpu-workflow}
fit <- cudacell_workflow(
  counts,
  n_hvg = 20,
  n_components = 5,
  k = 5,
  batch_size = 8,
  device = "cpu"
)

fit
```

Normalization remains sparse, while PCA scores and neighbour results retain
cell identifiers:

```{r cpu-identifiers}
class(fit$normalized)
identical(rownames(fit$pca$x), colnames(counts))
identical(rownames(fit$neighbors$index), colnames(counts))
```

## Read the provenance table

`cuda_provenance()` returns the ordered stages that produced an object. The
workflow table includes fixed CPU preprocessing and the delegated
`cudaverse` PCA and kNN stages:

```{r cpu-provenance}
cuda_provenance(normalized)
cuda_provenance(variable)
cuda_provenance(pca)
cuda_provenance(neighbors)
workflow_provenance <- cuda_provenance(fit)
workflow_provenance
attr(workflow_provenance, "compute_device")
```

The columns answer different questions:

| Column | Meaning |
|---|---|
| `requested_device` | What selected the stage: an explicit or automatic request, a fixed CPU stage, or an inherited choice. |
| `device` | Where that stage actually computed: `"cpu"` or `"cuda"`. |
| `backend` | The concrete implementation, such as `Matrix`, `base`, `stats`, or `torch`. |
| `selection_reason` | Why the stage used that device, including fixed CPU algorithms, input transfers, or CUDA availability. |
| `fallback` | `TRUE` only when an `"auto"` request selected CPU because CUDA was unavailable. |
| `output_device` | Where the stage result was materialized; public single-cell results are ordinary CPU-side R objects. |

The aggregate `compute_device` is `"cpu"` when all stages are CPU, `"cuda"`
when all stages compute on CUDA, and `"hybrid"` when a workflow contains both.
The returned object's CPU location does not imply fallback: a CUDA stage can
intentionally materialize its result for R and record `output_device = "cpu"`.

## Stage-by-stage backend semantics

| Function | Stage | Device/backend semantics |
|---|---|---|
| `cuda_normalize_counts()` | `normalization` | Fixed CPU `Matrix` sparse operations. |
| `cuda_hvg()` | `hvg` | Fixed CPU `Matrix` feature statistics and base-R ranking. |
| `cuda_cell_pca()` | `normalization`, `hvg` | Fixed CPU sparse preprocessing. |
| `cuda_cell_pca()` | `pca_preprocessing`, `pca_decomposition` | `stats` on CPU or `torch` on CUDA; the final R result is on CPU. |
| `cuda_cell_neighbors()` | inherited source stages | Preserves PCA or other input provenance. |
| `cuda_cell_neighbors()` | `knn_distance` | `base` on CPU or `torch` on CUDA. |
| `cuda_cell_neighbors()` | `knn_neighbor_selection` | Fixed CPU deterministic ordering and selection. |
| `cudacell_workflow()` | all stages above | CPU for sparse preprocessing; hybrid when PCA/kNN distance uses CUDA. |

An upstream `cudasparse` input contributes prefixed source stages. If a CUDA
sparse input must become a CPU `dgCMatrix`, provenance also records an
`input_materialization` transfer rather than hiding it.

## Runtime diagnostics and device requests

The shared diagnostic does not install or download torch:

```{r diagnostics}
diagnostics <- cudaverse::cuda_diagnostics()
diagnostics
```

The request contract is:

- `"cpu"` guarantees the portable CPU implementation.
- `"cuda"` is strict. An unavailable runtime raises
  `cudaverse_cuda_unavailable` before CPU preprocessing starts.
- `"auto"` selects CUDA when usable; otherwise affected stages use CPU and
  record `fallback = TRUE` with a stable reason.
- Fixed CPU stages such as normalization have `requested_device =
  "fixed-cpu"` and `fallback = FALSE`. They are part of the algorithm, not a
  failed GPU request.
- A CUDA execution error after selection is reported; the workflow does not
  retry silently on CPU.

## Optional CUDA workflow

The same input can exercise every CUDA-aware single-cell stage:

```{r optional-cuda}
if (isTRUE(diagnostics$cuda_available)) {
  cuda_fit <- cudacell_workflow(
    counts,
    n_hvg = 20,
    n_components = 5,
    k = 5,
    batch_size = 8,
    device = "cuda"
  )

  cuda_table <- cuda_provenance(cuda_fit)
  cuda_table
  attr(cuda_table, "compute_device")
} else {
  message("CUDA example skipped: ", diagnostics$reason)
}
```

A successful explicit CUDA workflow is still `hybrid`: normalization, HVG,
neighbour selection, and R result materialization remain on CPU, while PCA and
kNN distance stages run on CUDA.

## Dense PCA and kNN memory

Normalization and HVG calculations retain sparse `Matrix` storage.
`cuda_cell_pca()` then selects `n_hvg` rows and materializes a dense
cell-by-feature matrix before PCA. A double copy of that selected matrix alone
uses roughly `8 * cells * n_hvg` bytes; centring, scaling, decomposition, and
CPU/GPU copies need additional working memory. Lower `n_hvg` before lowering
`n_components` when dense input memory is the constraint, because the selected
matrix is allocated before decomposition.

Exact kNN remains quadratic in time, but it does not retain a complete
cell-by-cell distance matrix. The resident double distance block is
approximately `8 * min(batch_size, cells) * cells` bytes, and the returned
distances use about `8 * cells * k` bytes plus an integer index matrix. Lower
`batch_size` to reduce peak memory; doing so does not change the neighbours.

## Hardware-enforced parity gate

Normal R package checks must work without CUDA and may conditionally skip the
optional example above. The cudaverse hardware contract is separate: it runs
on a self-hosted runner labelled `cuda`, sets
`CUDAVERSE_REQUIRE_CUDA=true`, verifies `nvidia-smi`, requires a working torch
CUDA runtime and at least one device, and compares CPU with CUDA PCA, kNN, and
the complete single-cell workflow. It also asserts that CUDA stages are
present in provenance and the aggregate workflow is hybrid.

The same marker can enforce a local or downstream hardware job:

```{r hardware-gate}
require_cuda <- identical(
  tolower(Sys.getenv("CUDAVERSE_REQUIRE_CUDA", unset = "false")),
  "true"
)
if (require_cuda && !isTRUE(diagnostics$cuda_available)) {
  stop("CUDA hardware is required, but no usable device was detected.")
}
```

The package CUDA workflow runs manually, or automatically when the repository
variable `CUDAVERSE_NVIDIA_CI` is `enabled`. A required hardware job with
missing CUDA fails; it cannot pass by skipping GPU assertions.
