---
title: "A native SeuratObject v5 workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A native SeuratObject v5 workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`cudacell_seurat()` runs the complete cudacellr preprocessing, PCA, and exact
nearest-neighbour workflow while keeping every result in a native Seurat v5
subobject. The adapter requires only `SeuratObject >= 5.0.0`; installing the
full Seurat package is not necessary.

```{r install, eval=FALSE}
install.packages("SeuratObject")
```

## Start with a populated object

This example contains a second assay, cell metadata, identities, a reduction,
miscellaneous data, and a tool record. The workflow returns a modified copy
and retains all of them.

```{r fixture}
library(cudacellr)
library(Matrix)

set.seed(42)
counts <- Matrix(
  matrix(
    rpois(80 * 30, lambda = 2),
    nrow = 80,
    dimnames = list(
      paste0("gene_", seq_len(80)),
      paste0("cell_", seq_len(30))
    )
  ),
  sparse = TRUE
)
counts[1, ] <- counts[1, ] + 1

object <- SeuratObject::CreateSeuratObject(
  counts = counts,
  assay = "RNA",
  meta.data = data.frame(
    batch = rep(c("one", "two"), each = 15),
    row.names = colnames(counts)
  ),
  project = "preserved-project"
)
object[["ADT"]] <- SeuratObject::CreateAssay5Object(
  counts = counts[seq_len(5), , drop = FALSE]
)
object[["existing_pca"]] <- SeuratObject::CreateDimReducObject(
  embeddings = matrix(
    rnorm(ncol(counts) * 2),
    nrow = ncol(counts),
    dimnames = list(colnames(counts), c("OLDPC_1", "OLDPC_2"))
  ),
  assay = "RNA",
  key = "OLDPC_"
)
SeuratObject::Idents(object) <- factor(
  rep(c("A", "B"), each = 15)
)
SeuratObject::Misc(object, slot = "owner") <- "unchanged"
add_example_tool <- function(x) {
  SeuratObject::Tool(x) <- list(owner = "unchanged")
  x
}
object <- add_example_tool(object)
existing_tool_names <- SeuratObject::Tool(object)
```

## Run the workflow

Input assay and layer names are exact. `assay = NULL` uses the default assay,
while `layer = "counts"` is explicit by default.

```{r workflow}
result <- cudacell_seurat(
  object,
  assay = "RNA",
  layer = "counts",
  n_hvg = 30,
  n_components = 5,
  k = 5,
  batch_size = 8,
  device = "cpu"
)

SeuratObject::Assays(result)
SeuratObject::Reductions(result)
SeuratObject::Neighbors(result)
```

The default outputs use namespaced native locations:

```{r outputs}
normalized <- SeuratObject::LayerData(
  result[["CUDACELL"]],
  layer = "data",
  fast = FALSE
)
normalized[1:3, 1:3]

SeuratObject::Embeddings(result[["cudacell_pca"]])[1:3, , drop = FALSE]
SeuratObject::Indices(result[["cudacell_knn"]])[1:3, , drop = FALSE]
result[["CUDACELL"]][[]][
  1:3,
  c(
    "cudacell_mean",
    "cudacell_variance",
    "cudacell_dispersion",
    "cudacell_hvg",
    "cudacell_hvg_rank"
  )
]
```

`cudacell_hvg_rank` is the one-based global dispersion rank across every
input feature. `cudacell_hvg` marks the selected top `n_hvg` features, which
are also registered through `SeuratObject::VariableFeatures()`.

The exact directed kNN result is stored as a `Neighbor`, including both
integer indices and distances. The PCA `DimReduc` includes cell embeddings,
feature loadings, standard deviations, and the `CUDACELLPC_` key.

## Verify preservation

The input object is unchanged, and unrelated state survives in the result:

```{r preservation}
identical(
  SeuratObject::LayerData(object[["RNA"]], "counts"),
  SeuratObject::LayerData(result[["RNA"]], "counts")
)
identical(object[["ADT"]], result[["ADT"]])
identical(object[["existing_pca"]], result[["existing_pca"]])
identical(SeuratObject::Idents(object), SeuratObject::Idents(result))
identical(
  SeuratObject::Misc(object, slot = "owner"),
  SeuratObject::Misc(result, slot = "owner")
)
all(vapply(
  existing_tool_names,
  function(name) {
    identical(
      SeuratObject::Tool(object, slot = name),
      SeuratObject::Tool(result, slot = name)
    )
  },
  logical(1)
))
```

cudacellr adds its own deterministic `cudacell_seurat` tool record without
replacing existing tool records.

## Collision and realization safety

Every target assay, reduction, neighbour, feature-metadata column,
cell-metadata column, tool record, and Seurat key is checked before device
selection, layer access, realization, or numerical work. A second run with
the same names fails:

```{r collision, error=TRUE}
cudacell_seurat(
  result,
  n_hvg = 30,
  n_components = 5,
  k = 5,
  device = "cpu"
)
```

Use `overwrite = TRUE` only when replacing those exact cudacellr outputs is
intentional. Cross-type name or key collisions are never overwritten.

Only ordinary in-memory `matrix` and `Matrix` layers are accepted by default.
Delayed, on-disk, and other matrix-like layers are not silently collected.
Set `realize = TRUE` only after confirming that the selected layer fits in
memory; the original layer class and materialization decision are recorded.

For `device = "cuda"`, CUDA availability is validated before the layer is
read or realized. The request remains strict and never silently falls back to
CPU.

## Provenance

The native tool record contains identifiers, parameters, output locations,
and the same ordered `cudaverse-stage/1` compute provenance used by the
matrix and SingleCellExperiment workflows:

```{r provenance}
cuda_provenance(result)
SeuratObject::Tool(result, slot = "cudacell_seurat")$outputs
SeuratObject::Tool(result, slot = "cudacell_seurat")$parameters
```
