---
title: "A native SingleCellExperiment workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A native SingleCellExperiment workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`cudacell_sce()` runs the complete cudacellr preprocessing, PCA, and nearest
neighbour workflow while keeping results in the native
`SingleCellExperiment` data model. It returns a modified copy; the input
object is not changed.

The Bioconductor packages are optional so matrix-only users do not acquire a
large mandatory dependency tree:

```{r install, eval=FALSE}
BiocManager::install("SingleCellExperiment")
```

## Start with a populated object

This example intentionally includes an unrelated assay, row and column
metadata, a pre-existing reduced dimension, top-level metadata, size factors,
and an alternative experiment. All of them must survive the workflow.

```{r fixture}
library(cudacellr)
library(Matrix)
library(SingleCellExperiment)

set.seed(42)
counts <- matrix(
  rpois(80 * 30, lambda = 2),
  nrow = 80,
  ncol = 30,
  dimnames = list(
    paste0("gene_", seq_len(80)),
    paste0("cell_", seq_len(30))
  )
)
counts[1, ] <- counts[1, ] + 1
counts <- Matrix(counts, sparse = TRUE)

sce <- SingleCellExperiment(
  assays = list(
    counts = counts,
    untouched = counts * 2
  ),
  rowData = S4Vectors::DataFrame(
    symbol = paste0("symbol_", seq_len(nrow(counts)))
  ),
  colData = S4Vectors::DataFrame(
    batch = rep(c("one", "two"), each = 15)
  ),
  metadata = list(owner = "unchanged"),
  reducedDims = list(
    EXISTING = matrix(rnorm(60), nrow = 30)
  )
)
sizeFactors(sce) <- rep(1, ncol(sce))
altExp(sce, "spike") <- SingleCellExperiment(
  assays = list(
    counts = matrix(
      rpois(5 * ncol(sce), 1),
      nrow = 5,
      dimnames = list(paste0("spike_", 1:5), colnames(sce))
    )
  )
)
```

## Run the workflow

The default output names are namespaced so common pre-existing `logcounts` and
`PCA` entries are not overwritten:

```{r workflow}
result <- cudacell_sce(
  sce,
  assay = "counts",
  n_hvg = 30,
  n_components = 5,
  k = 5,
  batch_size = 8,
  device = "cpu"
)

assayNames(result)
reducedDimNames(result)
colPairNames(result)
```

The result locations follow their SingleCellExperiment semantics:

```{r outputs}
assay(result, "cudacell_logcounts")[1:3, 1:3]
reducedDim(result, "CUDACELL_PCA")[1:3, , drop = FALSE]
rowData(result)[
  1:3,
  c(
    "cudacell_mean",
    "cudacell_variance",
    "cudacell_dispersion",
    "cudacell_hvg",
    "cudacell_hvg_rank"
  )
]
colPair(result, "CUDACELL_KNN")
```

`CUDACELL_KNN` is a directed `SelfHits` object. Its metadata columns contain
distance and neighbour rank. Unlike an integer matrix hidden in top-level
metadata, this representation automatically remaps cell indices when the SCE
is subset.

## Verify preservation

The original object is unchanged and unrelated fields are retained:

```{r preservation}
identical(assay(sce, "counts"), assay(result, "counts"))
identical(assay(sce, "untouched"), assay(result, "untouched"))
identical(rowData(sce)$symbol, rowData(result)$symbol)
identical(colData(sce)$batch, colData(result)$batch)
identical(reducedDim(sce, "EXISTING"), reducedDim(result, "EXISTING"))
identical(altExp(sce, "spike"), altExp(result, "spike"))
identical(sizeFactors(sce), sizeFactors(result))
identical(metadata(sce)$owner, metadata(result)$owner)
```

Feature statistics are mapped by their original integer row position, not by
feature name. Duplicate feature names are therefore safe.

## Normalization and size factors

The default normalized assay uses natural `log1p`, so it is named
`cudacell_logcounts` rather than the Bioconductor convention `logcounts`,
which commonly implies log base 2. Set `log1p = FALSE` to create
`cudacell_normalized`, or choose an explicit output name:

```{r normalization, eval=FALSE}
result <- cudacell_sce(
  sce,
  log1p = FALSE,
  normalized_assay = "normalized_counts",
  scale_factor = 10000,
  n_hvg = 30,
  n_components = 5,
  k = 5,
  device = "cpu"
)
```

The actual divisor, `library_size / scale_factor`, is always stored in
`colData(result)$cudacell_size_factor`. Canonical `sizeFactors(result)` are
left untouched unless `set_size_factors = TRUE`. Replacing existing canonical
size factors additionally requires `overwrite = TRUE`.

## Collision and memory safety

Every target assay, reduced dimension, colPair, rowData column, colData
column, and the `cudacellr` metadata record is checked before computation.
Rerunning with the same names fails rather than discarding data:

```{r collision, error=TRUE}
cudacell_sce(
  result,
  n_hvg = 30,
  n_components = 5,
  k = 5,
  device = "cpu"
)
```

Use `overwrite = TRUE` only when replacing those exact cudacellr outputs is
intentional. No unrelated field is replaced.

On-disk and other delayed assays are also rejected by default. cudacellr does
not silently turn a potentially enormous delayed assay into an in-memory
matrix. `realize = TRUE` is an explicit opt-in after checking the expected
memory requirement.

## Provenance and downstream embeddings

The lightweight metadata record contains parameters, output locations, and
the same validated `cudaverse-stage/1` compute stages as the matrix workflow:

```{r provenance}
cuda_provenance(result)
metadata(result)$cudacellr$parameters
```

Per-feature, per-cell, and graph-shaped results do not live in top-level
metadata because that metadata is not remapped by row or column subsetting.
For a downstream embedding, pass `result` directly to a compatible
`cudaembedr` function or explicitly use
`reducedDim(result, "CUDACELL_PCA")`.
